---
title: "Explore GEO"
output: html_notebook
---

## Packages

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
if (!requireNamespace("GEOmetadb", quietly = TRUE))
  BiocManager::install("GEOmetadb")
if (!requireNamespace("DBI", quietly = TRUE))
  install.packages("DBI")
if (!requireNamespace("RSQLite", quietly = TRUE))
  install.packages("RSQLite")
if (!requireNamespace("dplyr", quietly = TRUE))
  install.packages("dplyr")
if (!requireNamespace("ggplot2", quietly = TRUE))
  install.packages("ggplot2")

library(GEOmetadb)
library(DBI)
library(RSQLite)
library(ggplot2)
library(dplyr)

if (!file.exists('~/projects/Megan_Ding/in_class_exercises/using_geometadb/GEOmetadb.sqlite')) getSQLiteFile()
file.info('~/projects/Megan_Ding/in_class_exercises/using_geometadb/GEOmetadb.sqlite')


con <- dbConnect(SQLite(), '~/projects/Megan_Ding/in_class_exercises/using_geometadb/GEOmetadb.sqlite')

dbListTables(con)

dbListFields(con, 'gse')

```


## Q1. What type of data is it? Describe it.
```{r}
vect2sql <- function(x){
  paste("('", paste(x, collapse = "','"), "')", 
                 sep = "")
}
gse_ids <- c("GSE201427","GSE166530","GSE162183","GSE186458")
gse_sql <- vect2sql(gse_ids)
# dbListTables(con)

# dbListFields(con, 'gpl')

sql <- paste(
  "SELECT DISTINCT gse_gpl.gse, gse_gpl.gpl, gse.title",
  "FROM gse JOIN gse_gpl ON gse_gpl.gse = gse.gse",
  "WHERE gse.gse IN", gse_sql, sep = " ")
series <- dbGetQuery(con, sql)
```

GSE186458, GSE166530, GSE162183: Illumina NovaSeq 6000 (Homo sapiens), high-throughput sequencing (RNA-seq)

GSE201427: Illumina HiSeq 4000 (Homo sapiens), high-throughput sequencing (RNA-seq)

## Q2. What disease or model is it using?
```{r}
sql <- paste(
  "SELECT DISTINCT gse, overall_design, variable, variable_description, pubmed_id, summary",
  "FROM gse",
  "WHERE gse IN", gse_sql, sep = " ")
model <- dbGetQuery(con, sql)
```

GSE162183: Single cell RNA-seq study of human psoriasis skin.
GSE166530: Differential expression analysis of RNA-Seq data from COVID-19 positive samples against COVID-19 negative control.
GSE186458: Genome-wide patterns of DNA methylation across a variety of cell types in healthy adults.
GSE201427: Panc1 cells were treated either with control siRNA (siCtrl) or siRNA targeting SF3B1 (siSF3B1). siCtrl cells were exposed to 1% O2 or 21% O2, siSF3B1 cells to 1% O2 for 8 hours. 1% O2 mimics hypoxia.

## Q3. How many samples are there? What type of data is associated with the record?
```{r}
like_sql <- paste(paste0("series_id LIKE '%", gse_ids, "%' OR"), collapse = " ")
like_sql <- substr(like_sql, 1, nchar(like_sql)-3)
sql <- paste(
  "SELECT series_id, gsm, title, type, source_name_ch1, characteristics_ch1, label_ch1, treatment_protocol_ch1, source_name_ch2, characteristics_ch2, label_ch2, treatment_protocol_ch2, description",
  "FROM gsm",
  "WHERE", like_sql, sep = " ")
	
all_gsm_attr <- dbGetQuery(con, sql)
table(all_gsm_attr$series_id)
```

GSE162183: 6 samples, scRNA-seq
GSE166530: 41 samples, bulk RNA-seq
GSE186458: 253, bulk RNA-seq
GSE201427: 8, bulk RNA-seq

## Q4. How many control and how many disease?
```{r}
all_gsm_attr_ctrl <- all_gsm_attr[grepl("control|normal|healthy|wild|ctrl|Ctrl|WT|Neg|neg", all_gsm_attr$source_name_ch1) | grepl("control|normal|healthy|wild|ctrl|Ctrl|WT|Neg|neg", all_gsm_attr$characteristics_ch1), ]

all_gsm_attr_disease <- all_gsm_attr[grepl("tumo?r|cancer|patient|disease|mutant|treated|stim|mut|KO|KD|Pos|pos", all_gsm_attr$source_name_ch1) | grepl("tumo?r|cancer|patient|disease|mutant|treated|stim|mut|KO|KD|Pos|pos", all_gsm_attr$characteristics_ch1), ]
table(all_gsm_attr_ctrl$series_id)
table(all_gsm_attr_disease$series_id)
table(all_gsm_attr$source_name_ch1[all_gsm_attr$series_id == "GSE186458"])
```

GSE162183: 3 control => 3 disease (since there are 6 total samples, but nothing matches my disease query)
GSE166530: 5 control, 36 disease
GSE186458: No case-control, but has several tissue sources as tabulated above.
GSE201427: 5 control, 3 treatment (since there are 8 total samples, but nothing matches my treatment query)

## Q5. Can you find out any information about the GEO dataset - GSE264108. Why or Why not?

```{r}
sql <- paste(
  "SELECT DISTINCT *",
  "FROM gse",
  "WHERE gse LIKE '%GSE264108%'", sep = " ")
dbGetQuery(con, sql)
sql <- paste(
  "SELECT DISTINCT *",
  "FROM gse_gpl",
  "WHERE gse LIKE '%GSE264108%'", sep = " ")
dbGetQuery(con, sql)

sql <- "SELECT submission_date FROM gse"
dates <- dbGetQuery(con, sql)
max(dates[[1]])
```

None of the tables have GSE264108 in them. This is because this GSE became public on Jan 25, 2025, but the oldest GSE in the database is from Feb 29, 2024. This gets at the issue with this method of querying GEO as we rely on a static instance of the database that may be almost 2 years out of date.

